HaploQA vs. R/qtl2 Haplotype Reconstructions

### 2022-12-22 draft_report.Rmd
12/22 update

@@

@@

#### 2.2. Data Retrieval

Input data was retrieved from haploqa.jax.org, and converted into the qtl2 input format. The data wrangling consists of two parts:
1. scrapes the HTML elements of the haploqa website and downloads the SNP level reports for each individual within each cross
2. concatenate the individuals within each cross into a single genotype file, converts the genotype file into the required input files for qtl2

However, haploqa.jax.org contains a large number of uploaded samples and the input data for each cross needs to be retrieved from their according sample. For example, for cross 'Collaborative Cross', the URL to the sample is http://haploqa-dev.jax.org/tag/Collaborative%20Cross.html

A configuration file, 'annotations_config.csv', is stored in the environment, which allows the pipeline to automatically obtain all relevant information associated with each cross such as the sample URL, input data directory, output data directory, marker annotation file names, etc. 
The purpose of this configuration file is to avoid having to pass a large list of parameters into each pipeline function.

@@

@@

There are three major pipelines in the repository:
a. 'haplotype_reconstruction_pipeline' - haplotype reconstructions for an entire cross (all individuals within the cross) 
b. 'sample_haplotype_reconstruction' - haplotype reconstructions for one individual 
c. 'truth_model_reconstruction' - haplotype reconstructions for optimal/truth models

@@

#### 2.3. Results

##### 2.3.1. GigaMUGA

@@

# remove DO from memory
```

@@

```{r}
f2_results <- haplotype_reconstruction_pipeline('F2', list_pheno, qtl2_file_gen = F, samples_gen = F)
```

@@

```{r}
# ### GigaMUGA mutant
gigamuga_url <- 'http://haploqa-dev.jax.org/tag/Jackson_Lab_Mahaffey_MURGIGV01_20160102_FinalReport.html'
url_domain <- 'http://haploqa-dev.jax.org/'
marker_type <- 'GigaMUGA'
html_table <- read_html(gigamuga_url) %>% html_nodes("a") %>% html_attr("href")
url_list <- paste0(url_domain, html_table[grepl('/sample', html_table)])
summary_df <- sample_summary_scrape(read_html(gigamuga_url), url_list, marker_type)
summary_df$`% No Call` <- as.numeric(gsub("%", "", summary_df$`% No Call`))
summary_df <- summary_df[summary_df$`% No Call` < 10,]

ind_gm <- summary_df[summary_df$`Haplotype Candidate` == 'False',]$`ID`
ind_gm <- ind_gm[!ind_gm %in% c('QU', 'SM', 'SX', 'T5', 'TG', 'TR', 'TS', 'TX', 'UF', 'UR', 'VA', 'VE', 'VG', 'VK', 'VN', 'W4', 'XS', 'Y8', 'Z8', 'ZG')] # screentime error
ind_gm <- ind_gm[!ind_gm %in% c('RY', 'S8', 'SG', 'SR', 'Y2', 'YA', 'YJ', 'YT')] # only one founder

gigamuga_results <- list()
gm_pct_res <- list()
for (ind in ind_gm) {
  print(ind)
  sample_gm_res <- sample_haplotype_reconstruction('MURGIGV01', ind, samples_gen = T, qtl2_file_gen = T)
  df <- ind_geno_comp(sample_gm_res, ind, 'MURGIGV01')

  gm_pct_res[[ind]] <- df
  gigamuga_results[[ind]] <- sample_gm_res
}
```

##### 2.3.2. MiniMUGA

@@

##### 2.3.3. Shiny Visualizations

@@

#### 3.Statistical Results

@@

>>>>

#### 2.1. Environment set-up

@@

There is no need to pre-download anything from any other websites or manually create any directories, as those processes were automated by the pipelines. However, the automation depend heavily on the CSV files stored in the directory, and without proper set-ups of the CSV files, the pipeline will not run. Please make sure the following CSVs are present in the directory after cloning the repo:

1. annotations_config.csv - configuration file containing all technical set-ups for the pipeline, details will be explained in the next section.

2. founder_lookup_table.csv - lookup table containing all possible geno codes (AA, BB, AB, AY, BY, etc.) and their according number codes (1, 2, 3, 4, etc.)

3. founder_strains_table.csv - lookup table containing the standard 8 founder strains (A/J, C57BL/6J, etc.) and their according letter codes (A, B, etc.) This is primarily used for CC and DO to ensure the 8 founders are encoded in the correct order.

@@

There are three major pipelines in the repository:

a. 'haplotype_reconstruction_pipeline' - haplotype reconstructions for an entire cross (all individuals within the cross)

b. 'sample_haplotype_reconstruction' - haplotype reconstructions for one individual

c. 'truth_model_reconstruction' - haplotype reconstructions for optimal/truth models

To run a cross, make sure to modify the configuration file, 'annotations_config.csv', that's stored in the environment. This file allows the pipeline to automatically obtain all relevant information associated with each cross such as the sample URL, input data directory, output data directory, marker annotation file names, etc. 
The purpose of this configuration file is to avoid having to pass a large list of parameters into each pipeline function.

@@

If the target cross exists in the config file, the pipeline will:

1. get the filename entered in the 'annot_file' column as the annotation file from 'https://raw.githubusercontent.com/kbroman/MUGAarrays/main/UWisc/'

2. get the input data files for all individuals in the cross from the HaploQA url entered in the 'url' column

3. create a folder in the working directory (if not already exists) with the name entered in 'data_dir' column and outputs the scraped input files to this directory

4. create a folder in the working directory (if not already exists) with the name entered in 'qtl2_dir' column to output all the files for qtl2 input

5. get the founder data from the url given in 'founder_url' column

6. run qtl2 with the given value in the 'ngen' column

7. identify the appropriate crosstype to get the optimal model for the given cross using 'truth_model'

8. identify the founders present in the cross. This should be left blank for the mutant crosses or any other crosses where the number of founders ia not consistent between individuals

9. use 'marker_type' to identify the conditions of certain if loops in the pipeline. Should be 'GigaMUGA' or 'MiniMUGA' for the current pipelines

If the cross that the user is running haplotype reconstructions for does not exist in the config file, add a new row and fill each column accordingly.

@@

From the above set up, the scripts will look for gm_uwisc_v2.csv in https://raw.githubusercontent.com/kbroman/MUGAarrays/main/UWisc/ as the annotation file. The raw input data files will be scraped from http://haploqa-dev.jax.org/tag/Collaborative%20Cross.html and these files will be stored in a folder, 'haploqa_cc_gm', in the same directory as the repo, for further wrangling. The wrangled data that's ready for qtl2 read_cross function will be stored in a folder 'cc_qtl2_genail' also in the same directory as the repo. qtl2 will be use ngen = 4 to run haplotype reconstructions. The optimal model of CC is 'risib8' in qtl2.

@@

#### 2.2. Results

##### 2.2.1. GigaMUGA

@@

# remove DO from memory, use gc()
```

@@

```{r}
f2_results <- haplotype_reconstruction_pipeline('F2', list_pheno, qtl2_file_gen = T, samples_gen = T)
```

@@

```{r}
gm_res_dir <- file.path(results_dir, 'RDS', 'gm_results.rds')
gm_pct_dir <- file.path(shiny_pct_fp, 'gm_pct_cpmp.csv')
# ### GigaMUGA mutant
gigamuga_url <- 'http://haploqa-dev.jax.org/tag/Jackson_Lab_Mahaffey_MURGIGV01_20160102_FinalReport.html'
url_domain <- 'http://haploqa-dev.jax.org/'
marker_type <- 'GigaMUGA'
html_table <- read_html(gigamuga_url) %>% html_nodes("a") %>% html_attr("href")
url_list <- paste0(url_domain, html_table[grepl('/sample', html_table)])
summary_df <- sample_summary_scrape(read_html(gigamuga_url), url_list, marker_type)
summary_df$`% No Call` <- as.numeric(gsub("%", "", summary_df$`% No Call`))
summary_df <- summary_df[summary_df$`% No Call` < 10,]

ind_gm <- summary_df[summary_df$`Haplotype Candidate` == 'False',]$`ID`
ind_gm <- ind_gm[!ind_gm %in% c('QU', 'SM', 'SX', 'T5', 'TG', 'TR', 'TS', 'TX', 'UF', 'UR', 'VA', 'VE', 'VG', 'VK', 'VN', 'W4', 'XS', 'Y8', 'Z8', 'ZG')] # screentime error
ind_gm <- ind_gm[!ind_gm %in% c('RY', 'S8', 'SG', 'SR', 'Y2', 'YA', 'YJ', 'YT', 'XZ', 'Y9')] # only one founder
ind_gm <- ind_gm[7:length(ind_gm)]
#gigamuga_results <- list()
#gm_pct_res <- list()
skipped_inds <- list()
for (ind in ind_gm) {
  print(ind)
  skip <- FALSE
  tryCatch(sample_gm_res <- sample_haplotype_reconstruction('MURGIGV01', ind, samples_gen = T, qtl2_file_gen = T), error = function(e) {skip <<- TRUE})
  if(skip) {
     skipped_inds <- c(skipped_inds, ind)
     next}
  #df <- ind_geno_comp(sample_gm_res, ind, 'MURGIGV01')

  #gm_pct_res[[ind]] <- df
  gigamuga_results[[ind]] <- sample_gm_res
}
gigamuga_results <- readRDS(gm_res_dir)
saveRDS(gigamuga_results, gm_res_dir)
#write.csv(gm_pct_res, file.path(shiny_pct_fp, 'gm_pct_cpmp.csv'), row.names = F)
```

##### 2.2.2. MiniMUGA

@@

##### 2.2.3. Shiny Visualizations

@@

#### 3. Statistical Results

@@

Known issues:
1. Screen time error
2. Error in dev.off() : 
  QuartzBitmap_Output - unable to open file '/Users/linc/.local/share/rstudio/notebooks/AFB3D6E3-draft_report/1/8779F85C5B04ECDA/cvdr4ccbjgppi_t/_rs_chunk_plot_010.png'
